abxD01 Paper

### 2015-03-09 PAPER.Rmd

Through our antibiotic perturbation experiments we identified a panel of bacteria with either a strong positive or negative relationship with subsequent levels of *C. difficile* colonization. A large number of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, other *Clostridia*, and *Alistipes* were found to have strong potential resistance abilities against *C. difficile*. Conversely, several members of the *Proteobacteria* phylum, including *Pseudomonas*, *Streptococcus*, and especially *Escherichia* have strong potential positive influences on *C. difficile* colonization. These results were supported by their significant correlation values across several antibiotic regimens. By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had relatively increased resistance to *C. difficile* colonization. Using this collective microbiota data spanning a wide range of colonization resistance levels, we next moved beyond correlation to prediction of *C. difficile* colonization based on the composition of the microbiota. Through modeling we identified bacteria within the community which contributed greatest to accurately predicting *C. difficile* levels. This is the first analysis measuring and predicting colonization resistance levels based on the microbiota structure.

Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance. Correlation analysis identified whole families with significant negative or positive relationships with *C. difficile*. However, correlation analysis alone was insuficient in identifying all informative bacteria. Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C. difficile* yet improved model accuracy. *Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (Figure 5bc). While these bacteria may not have important direct effects on *C. difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C. difficile*. Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C. difficile* observed. For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C. difficile*. The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C. difficile* (*Figure 5C*). However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C. difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces. Similarly, while having high amounts of *Porphyromonadaceae* OTU13 in many instances resulted in low *C. difficile* colonization, 3 ampicillin treated, 2 cefoperazone (both 0.1 mg/ml and 0.3 mg/ml) treated, and 2 streptomycin (0.1 mg/ml) treated microbiotas (OTU13 abundance at least ~5%), had levels of *C. difficile* ranging from 10<sup>2</sup> to upwards of 10<sup>8</sup>. [Can say a similar thing about *Porphyromonadaceae* OTU6] Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).

Furthermore, the ability of *C. difficile* to colonize a community does not appear to be dependent soley on the loss of protective species. <> Also having high levels of an OTU with high positive correlation with *C. difficile* is not a requirement for *C. difficile* colonization. For example,  *Escherichia* OTU11 was most commonly found among the most positively correlated OTUs with *C. difficile* across antibiotic treatments. However, following streptomycin treatment, these microbiotas harbored less than 0.5% relative abundance of OTU11 and were still colonized with as much as 10<sup>8</sup> *C. difficile* CFU/g feces.

>>>>

Through our antibiotic perturbation experiments we identified a panel of bacteria with either a strong positive or negative relationship with subsequent levels of *C. difficile* colonization. A large number of *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, other *Clostridia*, and *Alistipes* were found to have strong potential resistance abilities against *C. difficile*. Conversely, several members of the *Proteobacteria* phylum, including *Pseudomonas*, *Streptococcus*, and especially *Escherichia* have strong potential positive influences on *C. difficile* colonization. These results were supported by their significant correlation values across several antibiotic regimens. By either reducing the amount of microbiota perturbation or allowing recovery following perturbation, we were able to capture microbiota structures which had relatively increased resistance to *C. difficile* colonization. Using this collective microbiota data spanning a wide range of colonization resistance levels, we next moved beyond correlation to prediction of *C. difficile* colonization based on the composition of the microbiota. Through modeling we identified bacteria within the community which contributed greatest to accurately predicting *C. difficile* levels. This is the first analysis both measuring and predicting colonization resistance levels based on the microbiota structure.

Throughout our analysis, we observed examples of the significance of the microbiome as a whole over any individual bacterial species in contributing to resistance. Correlation analysis identified whole families with significant negative or positive relationships with *C. difficile*. However, correlation analysis alone was insuficient in identifying all informative bacteria. Random forest modeling allowed us to identify several bacterial OTUs that did not have significant correlations with *C. difficile* yet improved model accuracy. *Turicibacter* and *Akkermansia*, for example, both had non-significant correlations with *C .difficile* yet were more important for model accuracy over OTUs with stronger correlations (*Figure 5BC*). While these bacteria may not have important direct effects on *C. difficile*, they may interact positively or negatively with bacteria that do have strong relationships with *C. difficile*. Through the model analysis, we observed that no single bacteria was able to consistently explain the subsequent level of *C. difficile* observed. For example, the *Porphyromonadaceae* OTU3, along with the *Lachnospiraceae* OTU39, were the two most important OTUs in the model with strong negative correlations with *C. difficile*. The majority of microbiotas which had over 1% relative abundance of OTU3 had no detectable *C. difficile* (*Figure 5C*). However, a small group of microbiotas with greater than 1% OTU3 abundance, which included several streptomycin (0.1 mg/ml) treated, ampicillin treated (both 1 and 6 days recovery), and cefoperazone treated (both 0.1 mg/ml and 0.3 mg/ml) microbiotas, were not fully protected by the presence of this OTU alone and ranged in *C. difficile* levels from 10<sup>2</sup> to 10<sup>6</sup> CFU/g feces. Similarly, having low amounts of one "protective" bacteria such as *Lachnospriaceae* OTU39 does not automatically render the host susceptible to colonization. For instance streptomycin treated (0.1 mg/ml) mice did not harbor *Lachnospiraceae* OTU39, yet had undetectable levels of *C. difficile*. Thus the loss or absence of one protective bacterial species can be compensated by others, either related species or others with functional redundancy (as in redundancy in their protective abilities).

Furthermore, the dynamics of *C. difficile* colonization does not appear to be dependent soley on the loss of protective species. Both *Escherichia* OTU11 and *Streptococcus* OTU78 were positively correlated with *C. difficile* colonization and important in the random forest model. Notably, *Escherichia* OTU11 was among the strongest positively correlated OTUs with *C. difficile* across all antibiotic treatments, except for streptomycin. In fact, while streptomycin treatment (5 mg/ml) lead to high *C. difficile* colonization, these communities had very little, if any, *Escherichia* OTU11. This demostrates that while *Escherichia* may aid in colonization of *C. difficile*, it is not essential. These observations lead to a model of colonization by *C. difficile* in which antibiotic perturbation leads to loss of protective species.
